---
format: html
editor: 
  markdown: 
    wrap: sentence
bibliography: references.bib
---

```{r}
#| include: false

library(tidyverse)
```

-   Introduction for the themes and important stuff
-   Literature review is for expanding on the details

**TELL A STORY**

# Literature Review

## Introduction

An essential method in any form of data analysis is to visualize information.
Whether this is for exploring a data set or graphing statistical results, the output visualizations provide a deeper understanding of the underlying mechanisms.
For example, Anscombe's Quartet contains four bivariate data sets that all have, within small rounding errors, identical descriptive statistics (doi:10.2307/2682899).
While the trends of the data sets are apparent when graphed, they would be missed using the descriptive statistics (@fig-anscombe).
Furthermore, using data visualizations can help improve recognition of trends and recollection of information (sources).
**OTHER REFERENCES TO EXPAND:** @kosslyn1989**,** @meyer1997**,** @brewer2012**, @prasad2012**

```{r}
#| label: fig-anscombe
#| fig-cap: "Graphical display of Anscombe's Quartet. Each dataset has identical summary statistics: mu_x=9, mu_y= 7.50, sigma_x = 3.32, sigma_y = 2.03, rho_xy = 0.81"
anscombe.clean <- anscombe %>% 
  pivot_longer(1:8) %>% 
  separate(name, into = c('null', 'coord', 'set'), sep = '') %>% 
  select(-null) %>% 
  pivot_wider(names_from = coord, values_fn = list) %>% 
  unnest(c(x,y))

anscombe.clean %>% 
  group_by(set) %>% 
  summarize(x.mean = mean(x), x.sd = sd(x),
            y.mean = mean(y), y.sd = sd(y),
            corr = cor(x,y))

anscombe.clean %>% 
  ggplot(mapping = aes(x = x, y = y)) + 
  geom_point() + 
  geom_smooth(method = 'lm', se = F) +
  facet_wrap(~set, nrow = 1) + 
  theme_bw() + 
  theme(aspect.ratio = 1)
```

While many types of graphs exist, their existence does not imply that they are useful or even necessarily the best.
Early formal statistical testing of the effectiveness of graphs began around 100 years ago with an emphasis on numerical accuracy [@croxton1927; @croxton1932; @eells1926].
Recommendations from these studies suggested that graphs that produce better accuracy should be preferred over other types.
This practice continued as graphs were refined into their basic forms for evaluation.
A cornerstone study by @cleveland1984 focused on a concept referred to as *Elementary Perceptual Tasks (EPTs)*.
EPTs, as defined by Cleveland and McGill, are the action of extracting quantitative information from graphs, including position along a common scale, position along non-aligned scales, length, direction, angle, area, volume, curvature, and shading.
What they found when testing three of the EPTs is that conveying numeric information varied depending on how the data was presented, showing that graph types matter extracting information.
Further studies have replicated and expanded on the theory of EPTs and their role in producing better graphs [@heer2010].

## Data Visualization in Three Dimensions

There are many standard ways for handling univariate and bivariate data, but multivariate data becomes increasingly complex to visualize [@grinstein2001].
One method to visualize additional variables is to introduce color, shape, or size to convey a variable.
However, these methods tend to rank lower using the EPT theory for conveying quantitative information.
Another potential solution is to use facets to create separate graphs by one or more variables, but the number of graphs can become cumbersome rather quickly.

Consider the example shown in @fig-ept-examp.
Here, three continuous variables are shown and are generated to resemble a semi-sphere.
Using color, it is difficult to clearly establish the structure of z and its relation to x and y.
Studies have shown that color is often a poor indicator to compare numeric values (source).
Next, size is used to convey z, but the area of the dots makes comparisons of z across x and y difficult (area of circle to explain?).
Lastly, facets help to show some structure to z, but comparisons are now spread across multiple graphs and require more time to

```{r}
#| label: fig-ept-examp
#| layout-nrow: 2
#| fig-cap: "Three examples of the same data using different Elementary Perceptural Tasks to visualize three quantitative variables. The variables are generically named x, y, and z."
#| fig-subcap: 
#|  - "Graph using color to convey the value of z."
#|  - "Graph using size to convey the value of z."
#|  - "Graph using facets to convey groupings of z."
set.seed(5)
n = 100
x = runif(n, 0, 10); y = runif(n, 0, 10)
z = rnorm(n, 50, 20); z[z<0|z>100] <- 50
z = (x-5)^2 + (y-5)^2 + rnorm(n, 0, 1)
ept.examp <- cbind(x,y,z)

ept.examp %>% 
ggplot(mapping = aes(x = x, y = y, color = z)) + 
  geom_point() + 
  theme_bw() +
  theme(legend.position = 'top', aspect.ratio = 1)

ept.examp %>% 
ggplot(mapping = aes(x = x, y = y, size = z)) + 
  geom_point() + 
  theme_bw() +
  theme(legend.position = 'top', aspect.ratio = 1)

ept.examp %>% 
ggplot(mapping = aes(x = x, y = y)) + 
  geom_point() + 
  theme_bw() +
  theme(legend.position = 'top', aspect.ratio = 1/2) + 
  facet_wrap(~cut(z, breaks = 10*(-10:10)))

#Code to produce a 3d scatter plot of the data, not showing up in document rn so probably need to 
library(rgl)
plot3d(ept.examp)
view3d(theta = 25, phi = 20, zoom = 0.9)

```

In the case of three variables, it is natural to extend graphs into the native dimension of the data.
Three variables can each occupy a separate axis to uniquely identify a position in space.
The extra dimension is not without its additional challenges; perceiving information in three-dimensional space has further considerations, including viewing angle [@enns1990], shading of surfaces [@zacks1998], depth cues [@zacks; @fischer2000], and interactivity [@wang2022].

The current mantra is to avoid 3D graphs when possible (Many blog articles?).
This is somewhat justified in research directly comparing 2D and 3D graphs, although results are sometimes contradictory to the claim (Barfield (accuracy+confidence), Hughes (JND), Fisher (use), Zacks (depth cues)).
One interesting facet is that 3D graphs are not preferred when a suitable 2D alternative exists (fisher, at least one other too mentioned this).

One major limitation of studies surrounding 3D data visualizations is that they focus on 2D projections of the three dimensional space.
Granted, technology has advanced where it is now easier to truly represent 3D visualization spaces via 3D printing (Oswald) or virtual/augmented reality (Kraus).

````{=html}
<!-- Some examples of 2d plots, probably won't use table
```{r}
tribble(
  ~'Data', ~'Data Type', ~'Possible graphs',
  'Univariate', 'Categorical', 'Bar chart, pie chart',
  'Univariate', 'Numerical', 'Density chart, histogram',
  'Bivariate', 'Both categorical', 'Mosaic plot',
  'Bivariate', 'One categorical and one numerical', 'Box plot, bar chart',
  'Bivariate', 'Both numerical', 'Scatter plot, contour plot'
)
```
-->
````
